Use of general purpose graphics processing units with MODFLOW

### 2012-09-22 ResponseTo_GW20120518-0099_Review.tex

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim

\begin{document}

\noindent
\textbf{Use of general purpose graphics processing units with MODFLOW \\ (Ground Water Manuscript GW20120518-0099) }\\

@@

\noindent
\textbf{Authors:} Hughes, J.D., and White, J.T \\

\today \\

\noindent
We greatly appreciate the comments provided by the three journal reviewers and the editor. We have addressed all of the comments in this document. In general, we have made all of the suggested modifications. In a few cases, we disagreed with reviewer comments but have provided justification for all comments that were not accepted.

We have moved a significant amount of the GPGPU jargon to two sidebar boxes based on the editor's suggestion. We believe moving the jargon to sidebar boxes makes the manuscript more readable. The peer reviewer comment's were outstanding and if possible we would like to acknowledge them if they are willing to be identified.

Reviewer comments are included (\textit{in italics}) prior to our responses. Material revised based on journal reviewer comments are indicated using \color{blue} blue \color{black}text in the revised manuscript. \color{cyan} cyan \color{black}text in the revised manuscript represent changes made in response to internal USGS reviewers.

@@

We have also included our assessment of the best approach for being able to run a range of MODFLOW problems in parallel has been added to the \textbf{Conclusions} section.

\item \textit{Line 26 - ``can be *easily* parallelize"}

Suggested addition made.

\item \textit{Lines 28 - 34 - there is some time discrepancies here from 2005 to 1984, to 1988. Perhaps just say "MODFLOW" on line 28 and remove the ``since 1984" and insert ``numerous" before ``two-and" }

@@

We have added text to indicate that GPGPU parallelization has great potential for MODFLOW and other numerical simulators.

\item \textit{Line 213-214 - unclear. Consider rephrasing }

This sentence has been rephrased to clarify the description of the thickness of layer 1.

\item \textit{Line 351 - ``developed to *use* the PCG" (one does not "solve" the algorithm) }

### 2012-09-22 GW20120518-0099-r01.tex
Last revisions prior to submittal for BAO approval
Final revisions prior for resubmittal to the journal

\title{Use of general purpose graphical processing units with MODFLOW}
\author{Hughes, J.D.\\U.S. Geological Survey\\ Tampa, Florida\\ \texttt{jdhughes@usgs.gov} \and White, J.T. \\ U.S. Geological Survey\\ Tampa, Florida\\ \texttt{jwhite@usgs.gov}}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle
\begin{abstract}
To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed. The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete LU factorization, and generalized least-squares polynomial preconditioners. The UPCG solver also includes options for sequential and parallel solution on the central processing unit (CPU) using OpenMP. For simulations utilizing the GPGPU, all basic linear algebra operations are performed on the GPGPU; memory copies between the central processing unit CPU and GPCPU occur prior to the first iteration of the UPCG solver and after satisfying head and flow criteria or exceeding a maximum number of iterations. The efficiency of the UPCG solver for GPGPU and CPU solutions is benchmarked using simulations of a synthetic, heterogeneous  unconfined aquifer with tens of thousands to millions of active grid cells. Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW PCG solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize. Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be \color{blue}easily \color{black}parallelized.
\end{abstract}

@@

MODFLOW (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to \color{blue}numerous \color{black}two- and three-dimensional problems. Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively. \color{blue}Increased model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix. \color{black}To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers. The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (\color{cyan}SSOR\color{black}) linear solver. These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG2) linear solver (Hill 1990), \color{cyan}the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), \color{black}the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).

Parallelization of linear solvers is another approach for accommodating larger model sizes. \color{blue}Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem. For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem, the speedup would be 2, 3, 4, and 5, respectively. However, ideal parallelization is rarely, if ever, achieved because of communication overhead which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize. \color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}

@@

In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured preconditioned conjugate gradient (UPCG) linear solver for a unconfined groundwater flow problem. We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete LU (MILU0), and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU. And finally, we evaluate the performance of the UPCG solver on the GPGPU to parallel CPU simulations using the OpenMP parallel programming paradigm.

@@

For GPGPU code implementations, the concept is similar to OpenMP.  The main code is a sequential host code which forks the data into many threads at the invocation of a CUDA \color{blue}functions, called kernels (\textbf{Box 1})\color{black}.  \color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens-of-thousands of threads to execute the CUDA kernel function (\textbf{Box 2}). The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads. \color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} for MODFLOW and other numerical simulators that use Krylov subspace methods to solve large systems of simultaneous linear equations\color{black}.

@@

\noindent
where  $K_{xx}$, $K_{yy}$, $K_{zz}$ are the hydraulic conductivity [L/T] along the x, y, and z-coordinate axes which are assumed to be parallel to the major axes of hydraulic conductivity; $h$ is the groundwater head [L]; $W$ is a volumetric flux per unit volume [L$^{3}$/L$^{3}$T] representing sources/sinks of water; S$_{S}$ is the specific storage [L$^{-1}$]; and t is time [T].

@@

The conjugate gradient (CG) iterative method is an efficient method for solving large systems of linear equations having a square, symmetric, positive-definite coefficient matrix. Pseudocode for the preconditioned conjugate gradient method is given in \textbf{algorithm~\ref{pcgalg}}.

@@

All linear iteration steps, except for application of the zero fill-in incomplete LU (ILU0) and MILU0 preconditioners in line 3 of \textbf{algorithm~\ref{pcgalg}}, include operations that are independent for each active groundwater cell, indicating a potential for parallelization. In cases where (1) no preconditioner ($\mathbf{z = r}$) or (2) a Jacobi preconditioner ($\mathbf{M = I D}^{-1}$ -- where $\mathbf{I}$ is the identity matrix and $\mathbf{D}^{-1}$ is a matrix that only contains the inverse of the diagonal elements of $\mathbf{A}$), is selected, application of the preconditioner is highly parallelizable. For most practical problems, non-preconditioned and the Jacobi preconditioners will not significantly reduce the number of linear iterations ($i$) required to achieve convergence (Li and Saad 2010). As a result better preconditioners, such as incomplete factorizations (ILU0 or MILU0) or m-degree polynomial (GLSPOLY) approximations of $\mathbf{A}$, are generally needed. Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994). Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}

\subsection*{Numerical Implementation}
The UPCG linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations. Execution of the UPCG solver on the CPU, GPGPU, or combination of CPU and GPGPU is specified by the user in the UPCG input file.

Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row. This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS -- NVIDIA 2012). BLAS is a library of Basic Linear Algebra Subprograms (Lawson \textit{et al.} 1979)\color{black}. The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (\color{cyan}Saad \color{black}2003).

@@

\subsection*{Problem Description}
GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a 1,000 m square heterogeneous unconfined aquifer. The top and bottom of the aquifer are flat and specified to have elevations of 10 and -30 m, respectively. Models with horizontal discretizations of 200 $\times$ 200, 500 $\times$ 500, 1,000 $\times$ 1,000, 2,000 $\times$ 2,000, and 4,000 $\times$ 4,000 cells per layer were evaluated. Single layer models were evaluated for all horizontal discretizations and multi-layer models were evaluated for select horizontal discretizations to determine the effect that a 7-point stencil, and the resulting additional non-zero entries for each cell, has on GPGPU and parallel CPU execution time. Two- and three-layer models were evaluated for all horizontal discretizations except 4,000 $\times$ 4,000 cells per layer. Ten-layer models were also evaluated for horizontal discretizations ranging from 200 $\times$ 200 to 1,000 $\times$ 1,000 cells per layer. Additional horizontal and vertical discretizations were not evaluated because of GPGPU memory requirements for larger problems. For multi-layer models, a constant layer thickness was used for all but the first layer and was calculated by dividing 30 m by the number of layers. \color{blue}Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.\color{black}

@@

Hydraulic conductivity was not varied between layers in the multi-layer models evaluated. For models having less than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data was averaged for each base model cell contained in a coarse model cell. For models having more than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data was directly assigned to each higher-resolution cell contained within a coarse base model cell.

Head-dependent (general head) boundaries were specified on the left and right sides \color{cyan} in all layers \color{black} and no flow boundaries at the top and bottom. Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain \color{blue}$\left [ 10. / 1000. = 0.01 \text{ m/m} \right]$ \color{black}and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively. For the one layer 1,000 $\times$ 1,000 cell model, conductance and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively. The head-dependent boundaries cause an ambient groundwater flow from left to right. Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d. For multi-layer simulations, the pumping wells withdraw water from the lower model layer. There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries \color{cyan}and represents 46\% of the flow entering the model\color{black}. Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.

@@

All of the models simulated steady-state conditions. As a result, specific storage and specific yield values were not specified for the aquifer. Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads. Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels. \color{cyan}Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$RCLOSE m.\color{black}

@@

The UPCG solver was developed to \color{blue}use \color{black} the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners. An unstructured CSR format was used in the UPCG solver to facilitate use of the CUBLAS library. The UPCG solver also includes options for running sequential and parallel simulations on the CPU. Parallel CPU simulations are executed using the OpenMP programming paradigm.

A number of test problems were evaluated and indicate that it is possible to realize notable speedup through use of GPGPUs with MODFLOW.  CPU/GPGPU speedups ranging from 1.5 to 35 were observed for the Jacobi, MILU0, and GLSPOLY preconditioners when GPGPU runtimes were compared to CPU runtimes for the same preconditioner. Exceptions to notable CPU/GPGPU speedups were observed for the smallest problems where CPU-GPGPU memory transfer times were significant compared to solution time. CPU/GPGPU speedups with the MILU0 preconditioner was the smallest of all of the preconditioners (ranging from 0.99 to 1.936) because of the additional CPU-GPGPU memory transfers required during each inner (linear) iteration to apply the preconditioner.

@@

Source code for MODFLOW (MODFLOW--2005 version) with the UPCG solver is available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}. Development and compilation of CUDA and CUBLAS functions requires installation of the CUDA toolkit which includes the necessary CUDA drivers. NVIDIA also provides GPGPU code samples in the CUDA software development kit (SDK). Current versions of each of these are available free of charge from NVIDIA at \url{http://www.nvidia.com/content/cuda/cuda-downloads.html}. A 64-bit version of the MODFLOW--2005 executable with the UPCG solver and the linked OpenMP dynamic-link library (DLL) for the Windows 7 operating system and the Tesla C2050 GPGPU is also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg} and only requires installation of the CUDA drivers, which are also available as a stand-alone installation from NVIDIA, and a GPGPU with NVIDIA Compute Capability 2.0 or greater. This executable was linked against version 4.1.28 of the 64-bit CUDA toolkit. It is possible that the software may need to be modified for application of the code on GPGPU models other than the Tesla C2050. And finally, UPCG solver input instructions and python scripts used to create the MODFLOW datasets evaluated using the standard PCG solver with the MIC preconditioner and the UPCG solver are also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.

The USGS provides no warranty, expressed or implied, as to the correctness of the furnished software or the suitability for any purpose. The software has been tested, but as with any complex software, there could be undetected errors. Users who find errors are requested to report them to the USGS. References to non-USGS products, trade names, and (or) services are provided for information purposes only and do not constitute endorsement or warranty, express or implied, by the USGS, U.S. Department of Interior, or U.S. Government, as to their suitability, content, usefulness, functioning, completeness, or accuracy.\color{black}

@@

To use GPGPUs, specialized software must be developed to use a precompiled set of functions and routines which serve as a communication mechanism between a CPU and one or more GPGPU and pass instructions to the GPGPU.  These precompiled functions and routines are typically bundled into a code library that is accessed using a application programming interface (API).

\hspace{15pt}Kernels are blocks of code which are executed only on the GPGPU.  When the code is compiled, the compiler is instructed to use the API functions and routines to resolve the kernel block of code into GPGPU instructions. The kernels executed by the UPCG solver on the GPGPU are the CUDA-enabled BLAS operations in the preconditioned conjugate gradient algorithm.\\ \color{black}}}
\bigskip

@@

\hspace{15pt}Each kernel launches a CUDA grid, which is composed of one or more blocks of grouped threads. Threads within the same block can share certain types of physical memory on the GPGPU.  The number of blocks in a grid is determined by the size of the kernel data being processed or the number of SM. The total number of threads in a grid is equal to the number of threads per block times the number of blocks. CCs on each SM can run one or more blocks but blocks are allocated to a single CC. Threads within a block are further subdivided into warps for processing; a warp is the maximum number of threads that are managed and processed simultaneously within each block. Multiple warps execute one after another to complete block processing of the kernel. GPGPU architecture and kernel execution are shown graphically in \textbf{figure~\ref{FigParallel}}.

>>>>

\title{Use of general purpose graphics processing units with MODFLOW}
\author{Hughes, J.D. \and White, J.T.}
\date{}

\begin{document}
\maketitle
\begin{abstract}
To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed. The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete \color{cyan}lower-upper (LU) \color{black}factorization, and generalized least-squares polynomial preconditioners. The UPCG solver also includes options for sequential and parallel solution on the central processing unit (CPU) using OpenMP. For simulations utilizing the GPGPU, all basic linear algebra operations are performed on the GPGPU; memory copies between the central processing unit CPU and GPCPU occur prior to the first iteration of the UPCG solver and after satisfying head and flow criteria or exceeding a maximum number of iterations. The efficiency of the UPCG solver for GPGPU and CPU solutions is benchmarked using simulations of a synthetic, heterogeneous  unconfined aquifer with tens of thousands to millions of active grid cells. Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW \color{cyan}preconditioned conjugate gradient (PCG) \color{black}solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize. Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be \color{blue}easily \color{black}parallelized.
\end{abstract}

@@

MODFLOW (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to \color{blue}numerous \color{black}two- and three-dimensional problems. Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively. \color{blue}Increased model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix. \color{black}To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers. The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (\color{cyan}SSOR\color{black}) linear solver. These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG) linear solver (Hill 1990), \color{cyan}the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), \color{black}the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).

Parallelization of linear solvers is another approach for accommodating larger model sizes. \color{blue}Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem. For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem, the speedup would be 2, 3, 4, and 5, respectively. However, ideal parallelization is rarely, if ever, achieved because of communication overhead, which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize. \color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}

@@

In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured preconditioned conjugate gradient (UPCG) linear solver for a unconfined groundwater flow problem. We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete  \color{cyan}lower-upper (MILU0) factorization \color{black}, and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU. And finally, we evaluate the performance of the UPCG solver on the GPGPU to parallel CPU simulations using the OpenMP parallel programming paradigm.

@@

For GPGPU code implementations, the concept is similar to OpenMP.  The main code is a sequential host code, which forks the data into many threads at the invocation of a CUDA \color{blue}functions, called kernels (\textbf{Box 1})\color{black}.  \color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens of thousands of threads to execute the CUDA kernel function (\textbf{Box 2}). The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads. \color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} for MODFLOW and other numerical simulators that use Krylov subspace methods to solve large systems of simultaneous linear equations\color{black}.

@@

\noindent
where  $K_{xx}$, $K_{yy}$, $K_{zz}$ are the hydraulic conductivity [L/T] along the $x$, $y$, and $z$-coordinate axes, which are assumed to be parallel to the major axes of hydraulic conductivity; $h$ is the groundwater head [L]; $W$ is a volumetric flux per unit volume [L$^{3}$/L$^{3}$T] representing sources/sinks of water; S$_{S}$ is the specific storage [L$^{-1}$]; and $t$ is time [T].

@@

The conjugate gradient (CG) iterative method is an efficient method for solving large systems of linear equations having a square, symmetric, positive-definite coefficient matrix. Pseudocode for the PCG method is given in \textbf{algorithm~\ref{pcgalg}}.

@@

All linear iteration steps, except for application of the zero fill-in incomplete LU (ILU0) and MILU0 preconditioners in line 3 of \textbf{algorithm~\ref{pcgalg}}, include operations that are independent for each active groundwater cell, indicating a potential for parallelization. In cases where (1) no preconditioner ($\mathbf{z = r}$) or (2) a Jacobi preconditioner ($\mathbf{M = I D}^{-1}$ -- where $\mathbf{I}$ is the identity matrix and $\mathbf{D}^{-1}$ is a matrix that only contains the inverse of the diagonal elements of $\mathbf{A}$), is selected, application of the preconditioner is highly parallelizable. For most practical problems, non-preconditioned and the Jacobi preconditioners will not significantly reduce the number of linear iterations ($i$) required to achieve convergence (Li and Saad 2010). As a result better preconditioners, such as incomplete factorizations (ILU0 or MILU0) or m-degree polynomial (GLSPOLY) approximations of $\mathbf{A}$, are generally needed. Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994). Preconditioning using polynomial preconditioners, however, is simple, efficient, and easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}

\subsection*{Numerical Implementation}
The \color{cyan}unstructured preconditioned conjugate gradient (UPCG) \color{black}linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations. Execution of the UPCG solver on the CPU, GPGPU, or combination of CPU and GPGPU is specified by the user in the UPCG input file.

Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row (CSR) storage format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row. This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS -- NVIDIA 2012). BLAS is a library of Basic Linear Algebra Subprograms (Lawson \textit{et al.} 1979)\color{black}. The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (\color{cyan}Saad \color{black}2003).

@@

\subsection*{Problem Description}
GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a \color{cyan}1,000 m $\times$ 1,000 m \color{black} heterogeneous unconfined aquifer. The top and bottom of the aquifer are flat and specified to have elevations of 10 and -30 m, respectively. Models with horizontal discretizations of 200 $\times$ 200, 500 $\times$ 500, 1,000 $\times$ 1,000, 2,000 $\times$ 2,000, and 4,000 $\times$ 4,000 cells per layer were evaluated. Single layer models were evaluated for all horizontal discretizations and multi-layer models were evaluated for select horizontal discretizations to determine the effect that a 7-point stencil, and the resulting additional non-zero entries for each cell, has on GPGPU and parallel CPU execution time. Two- and three-layer models were evaluated for all horizontal discretizations except 4,000 $\times$ 4,000 cells per layer. Ten-layer models were also evaluated for horizontal discretizations ranging from 200 $\times$ 200 to 1,000 $\times$ 1,000 cells per layer. Additional horizontal and vertical discretizations were not evaluated because of GPGPU memory requirements for larger problems. For multi-layer models, a constant layer thickness was used for all but the first layer and was calculated by dividing 30 m by the number of layers. \color{blue}Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.\color{black}

@@

Hydraulic conductivity was not varied between layers in the multi-layer models evaluated. For models having less than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data were averaged for each base model cell contained in a coarse model cell. For models having more than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data were directly assigned to each higher-resolution cell contained within a coarse base model cell.

Head-dependent (general head) boundaries were specified on the left and right sides \color{cyan} in all layers \color{black} and no-flow boundaries at the top and bottom. Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain \color{blue}$\left [ 10. / 1000. = 0.01 \text{ m/m} \right]$ \color{black}and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively. For the one layer 1,000 $\times$ 1,000 cell model, conductances and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively. The head-dependent boundaries cause an ambient groundwater flow from left to right. Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d. For multi-layer simulations, the pumping wells withdraw water from the lower model layer. There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries \color{cyan}and represents 46\% of the flow entering the model\color{black}. Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.

@@

All of the models simulated steady-state conditions. As a result, specific storage and specific yield values were not specified for the aquifer. Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads. Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels. \color{cyan}Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$HCLOSE m.\color{black}

@@

The UPCG solver was developed to \color{blue}use \color{black} the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners. An unstructured compressed sparse row (CSR) storage format was used in the UPCG solver to facilitate use of the CUBLAS library. The UPCG solver also includes options for running sequential and parallel simulations on the CPU. Parallel CPU simulations are executed using the OpenMP programming paradigm.

Several test problems were evaluated and indicate that it is possible to realize notable speedup through use of GPGPUs with MODFLOW.  CPU/GPGPU speedups ranging from 1.5 to 35 were observed for the Jacobi, MILU0, and GLSPOLY preconditioners when GPGPU runtimes were compared with CPU runtimes for the same preconditioner. Exceptions to notable CPU/GPGPU speedups were observed for the smallest problems where CPU-GPGPU memory transfer times were significant compared to solution time. CPU/GPGPU speedups with the MILU0 preconditioner was the smallest of all of the preconditioners (ranging from 0.99 to 1.936) because of the additional CPU-GPGPU memory transfers required during each inner (linear) iteration to apply the preconditioner.

@@

Source code for MODFLOW (MODFLOW--2005 version) with the UPCG solver is available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}. Development and compilation of CUDA and CUBLAS functions requires installation of the CUDA toolkit, which includes the necessary CUDA drivers. NVIDIA also provides GPGPU code samples in the CUDA software development kit (SDK). Current versions of each of these are available free of charge from NVIDIA at \url{http://www.nvidia.com/content/cuda/cuda-downloads.html}. A 64-bit version of the MODFLOW--2005 executable with the UPCG solver and the linked OpenMP dynamic-link library (DLL) for the Windows 7 operating system and the Tesla C2050 GPGPU is also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg} and only requires installation of the CUDA drivers, which are also available as a stand-alone installation from NVIDIA, and a GPGPU with NVIDIA Compute Capability 2.0 or greater. This executable was linked against version 4.1.28 of the 64-bit CUDA toolkit. It is possible that the software may need to be modified for application of the code on GPGPU models other than the Tesla C2050. And finally, UPCG solver input instructions and python scripts used to create the MODFLOW datasets evaluated using the standard PCG solver with the MIC preconditioner and the UPCG solver are also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.

Although this software has been used by the U.S. Geological Survey (USGS), no warranty, expressed or implied, is made by the USGS or the U.S. Government as to the accuracy and functioning of the software program and related program material nor shall the fact of distribution constitute any such warranty, and no responsibility is assumed by the USGS in connection therewith.  Although the software has been tested, there could be undetected errors. Users are encouraged to report any errors to these authors.  Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.\color{black}

@@

To use GPGPUs, specialized software must be developed to use a precompiled set of functions and routines that serve as a communication mechanism between a CPU and one or more GPGPU and pass instructions to the GPGPU.  These precompiled functions and routines are typically bundled into a code library that is accessed using a application programming interface (API).

\hspace{15pt}Kernels are blocks of code that are executed only on the GPGPU.  When the code is compiled, the compiler is instructed to use the API functions and routines to resolve the kernel block of code into GPGPU instructions. The kernels executed by the UPCG solver on the GPGPU are the CUDA-enabled BLAS operations in the preconditioned conjugate gradient algorithm.\\ \color{black}}}
\bigskip

@@

\hspace{15pt}Each kernel launches a CUDA grid, which is composed of one or more blocks of grouped threads. Threads within the same block share physical memory on the GPGPU.  The number of blocks in a grid is determined by the size of the kernel data being processed or the number of SM. The total number of threads in a grid is equal to the number of threads per block times the number of blocks. CCs on each SM can run one or more blocks but blocks are allocated to a single CC. Threads within a block are further subdivided into warps for processing; a warp is the maximum number of threads that are managed and processed simultaneously within each block. Multiple warps execute one after another to complete block processing of the kernel. GPGPU architecture and kernel execution are shown graphically in \textbf{figure~\ref{FigParallel}}.
